---
title: "Visualizing Longitudinal Trajectories with ggkodom"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing Longitudinal Trajectories with ggkodom}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%"
)
has_lme4 <- requireNamespace("lme4", quietly = TRUE)
has_fdapace <- requireNamespace("fdapace", quietly = TRUE)
has_mgcv <- requireNamespace("mgcv", quietly = TRUE)
```

**ggkodom** is a ggplot2 toolkit for visualizing individual longitudinal
trajectories. Each subject gets its own visual lane — colored by
measurement value — so you can see patterns across subjects and time
at a glance. The package provides four display modes (swimlane, state-ribbon,
heatmap, circular), a one-call smoother, and FPCA-based decomposition tools.

## The data format

Every function in **ggkodom** expects a **long-format** data frame with at
least three columns: one for the subject identifier, one for time, and one for
the measurement value.

We will use a synthetic dataset that mimics repeated HbA1c measurements in a
diabetes monitoring study: 60 subjects, 4–8 irregularly-spaced visits,
sparse observations typical of electronic health records.

```{r generate-data}
library(ggkodom)

set.seed(1234)
n_subj <- 60
ids <- seq_len(n_subj)

make_subject <- function(i) {
  n_obs <- sample(4:8, 1)
  baseline <- rnorm(1, mean = 7.2, sd = 1.2)
  slope <- rnorm(1, mean = -0.05, sd = 0.15)
  times <- sort(sample(0:24, n_obs, replace = FALSE)) # months 0–24, no ties
  values <- baseline + slope * times + rnorm(n_obs, 0, 0.5)
  data.frame(id = paste0("S", i), time = times, hba1c = pmax(4, pmin(12, values)),
             stringsAsFactors = FALSE)
}

df <- do.call(rbind, lapply(ids, make_subject))
head(df, 8)
```

## 1. The classic swimlane

`kodom_swimlane()` produces one horizontal lane per subject. Segments are
colored by the measurement value using the Kadam-flower palette
(teal → gold → red). Subjects are sorted by their first observed value by
default, so the plot shows a natural gradient from healthiest (bottom) to
least controlled (top).

```{r swimlane-basic, fig.height = 5}
kodom_swimlane(df,
  id = "id",
  time = "time",
  value = "hba1c",
  color_breaks = c(5.7, 6.5, 8),
  n_sample = 40
)
```

The `color_breaks` argument anchors the three palette colors to clinically
meaningful cut-points (normal / prediabetes / diabetes / poor control).

### Tile (heatmap) layout

For datasets with many subjects or densely sampled time, the tile variant
bins time and fills each cell with an aggregate:

```{r swimlane-tile, fig.height = 5}
kodom_swimlane(df,
  id = "id",
  time = "time",
  value = "hba1c",
  geom = "tile",
  time_bins = 8,
  agg_fun = "mean",
  color_breaks = c(5.7, 6.5, 8),
  n_sample = 40
)
```

### Circular (Kodom) layout

The eponymous circular variant arranges subjects as radial spokes, time
increasing outward. The pattern resembles the Kadam flower
(*Neolamarckia cadamba*):

```{r swimlane-polar, fig.height = 5, fig.width = 5}
kodom_swimlane(df,
  id = "id",
  time = "time",
  value = "hba1c",
  coord = "polar",
  color_breaks = c(5.7, 6.5, 8),
  n_sample = 30
)
```

## 2. State-ribbon chart

When clinical thresholds matter more than exact values, `kodom_state()`
replaces the continuous gradient with discrete color bands:

```{r state-ribbon, fig.height = 5}
kodom_state(df,
  id = "id",
  time = "time",
  value = "hba1c",
  thresholds = c(5.7, 6.5, 8),
  state_labels = c("Normal", "Pre-DM", "DM", "Poor control"),
  n_sample = 40
)
```

The eye now catches **state transitions** — a subject whose lane shifts
from red to teal is improving — without being distracted by exact values.

## 3. Sorting lanes by an external key

Any of the built-in string options (`"first"`, `"last"`, `"mean"`,
`"median"`) can be passed to `sort_by`. You can also pass a named numeric
vector to impose an arbitrary order:

```{r sort-custom}
# Order by mean HbA1c, descending
mean_hba1c <- tapply(df$hba1c, df$id, mean)
# setNames creates the named vector kodom_swimlane expects
key <- sort(mean_hba1c, decreasing = TRUE)

kodom_swimlane(df,
  id = "id",
  time = "time",
  value = "hba1c",
  sort_by = key,
  color_breaks = c(5.7, 6.5, 8),
  n_sample = 40
)
```

## 4. Fitting a smoothed model with `kodom_smooth()` + `kodom_fit()`

`kodom_smooth()` wraps three smoothers — mixed models (`lme4`), GAMs
(`mgcv`), and functional PCA (`fdapace`) — behind a single call. The
returned list contains the fitted object and a `refit_fn` closure for
bootstrap refitting.

`kodom_fit()` then visualizes the **fitted curves** (not raw points) as a
swimlane — useful for comparing individual trajectories after accounting for
measurement noise.

```{r smooth-lmer, eval = has_lme4}
sm_lmer <- suppressMessages(
  kodom_smooth(df,
    method = "lmer",
    id = "id", time = "time", value = "hba1c"
  )
)

kodom_fit(sm_lmer$fit,
  id = "id",
  time = "time",
  color_breaks = c(5.7, 6.5, 8),
  n_sample = 40
)
```

```{r smooth-lmer-note, eval = !has_lme4, echo = FALSE}
message("Install 'lme4' to run this section.")
```

The fitted lines are smoother than the raw observations and reveal the
underlying trend for each subject.

## 5. FPCA decomposition

Functional PCA decomposes trajectories into a mean function and a small
number of orthogonal modes of variation (eigenfunctions). This is
especially useful when measurements are sparse and irregularly timed.

### Fitting FPCA

```{r fpca-fit, eval = has_fdapace}
sm_fp <- kodom_smooth(df,
  method = "fpca",
  id = "id", time = "time", value = "hba1c"
)
fp <- sm_fp$fit
```

```{r fpca-note, eval = !has_fdapace, echo = FALSE}
message("Install 'fdapace' to run the FPCA sections.")
```

### Eigenfunction plot (`kodom_components`)

`kodom_components()` shows the leading eigenfunctions — the directions in
function space that account for the most inter-subject variability:

```{r fpca-components, eval = has_fdapace, fig.height = 5}
kodom_components(fp, show_mean = TRUE)
```

Each panel is one functional principal component (FPC). A large positive
score on PC 1 means the subject's trajectory is displaced in the direction
of that curve relative to the mean.

### Perturbation plot (`kodom_perturbation`)

The Ramsay–Silverman perturbation plot overlays
$\mu(t) \pm 2\sqrt{\lambda_k}\,\phi_k(t)$ to show what each PC
*physically does* to the mean curve:

```{r fpca-perturbation, eval = has_fdapace, fig.height = 5}
kodom_perturbation(fp, K = 2)
```

A subject two standard deviations above average on PC 1 would have
a trajectory close to the upper (red) curve.

### Fitted reconstruction swimlane

`kodom_fit()` also works on FPCA objects:

```{r fpca-fit-plot, eval = has_fdapace, fig.height = 5}
ids_fp <- names(fp$inputData$Ly)[seq_len(nrow(fp$xiEst))]

kodom_fit(fp,
  ids = ids_fp,
  data = df,
  id = "id",
  time = "time",
  value = "hba1c",
  color_breaks = c(5.7, 6.5, 8),
  n_sample = 40
)
```

## 6. Score scatter and clustering

### Score scatter plot (`kodom_scores`)

`kodom_scores()` plots subjects in the two-dimensional FPC score space.
Each point is a subject; its position summarizes the shape of their
trajectory:

```{r fpca-scores, eval = has_fdapace}
kodom_scores(fp)
```

Color subjects by a continuous outcome (e.g., their mean HbA1c):

```{r fpca-scores-colored, eval = has_fdapace}
mean_val <- tapply(df$hba1c, df$id, mean)
# align to the IDs in the fit
ids_fp <- names(fp$inputData$Ly)[seq_len(nrow(fp$xiEst))]
groups <- as.numeric(mean_val[ids_fp])

kodom_scores(fp, groups = groups)
```

### Clustering in score space (`cluster_scores`)

`cluster_scores()` runs k-means (or Ward hierarchical) clustering on the
FPC scores and returns a named factor you can pipe directly into other
ggkodom functions:

```{r cluster, eval = has_fdapace}
cl <- cluster_scores(fp, k = 3, seed = 7)
table(cl)
```

Overlay the clusters onto the score scatter:

```{r cluster-scores, eval = has_fdapace}
kodom_scores(fp, groups = cl[ids_fp])
```

### Re-sort the swimlane by PC1 score

`kodom_sort_scores()` builds a named sort key from a chosen FPC:

```{r cluster-swimlane, eval = has_fdapace, fig.height = 5}
sort_key <- kodom_sort_scores(fp, pc = 1, ids = ids_fp)

df_cl <- df
df_cl$cluster <- cl[as.character(df_cl$id)]

kodom_swimlane(df_cl,
  id           = "id",
  time         = "time",
  value        = "hba1c",
  sort_by      = sort_key,
  facet_rows   = "cluster",
  color_breaks = c(5.7, 6.5, 8),
  n_sample     = 60
)
```

Faceting by cluster and ordering lanes by PC1 score reveals whether the
functional clusters correspond to distinct trajectory shapes — a quick
sanity-check before reporting phenotypes.

## Further customization

All ggkodom plots return a standard `ggplot` object and can be further
modified with any ggplot2 layer:

```{r customize}
library(ggplot2)

kodom_swimlane(df,
  id = "id",
  time = "time",
  value = "hba1c",
  color_breaks = c(5.7, 6.5, 8),
  n_sample = 40
) +
  labs(
    title = "HbA1c trajectories (n = 40)",
    subtitle = "Sorted by first observed value",
    x = "Months from enrolment",
    color = "HbA1c (%)"
  ) +
  theme(plot.title = element_text(face = "bold"))
```
