---
title: "Heatmap and Circular Variants: geom_kodom_heatmap and geom_kodom_circular"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Heatmap and Circular Variants: geom_kodom_heatmap and geom_kodom_circular}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 10,
  fig.height = 6,
  dpi = 300,
  out.width = "100%"
)
library(ggkodom)
library(ggplot2)
```

This vignette covers `geom_kodom_heatmap()` and `geom_kodom_circular()` — two
layouts built on the same aesthetic contract as `geom_kodom_line()` (see the
companion vignette for path-specific features: lane ordering, point shapes,
independent `size`/`linewidth`, etc.).

The guiding principle: use the **heatmap** when the cohort is large or when
per-observation timing matters less than overall trend; use the **circular**
("Kodom") layout when you want all subjects visible simultaneously in a
compact flower-like display.

## Sample data

Two datasets are used. The **small cohort** (25 patients, irregular visits)
is identical to the companion vignette and serves the circular examples. The
**large cohort** (100 patients) demonstrates where the heatmap layout
outperforms individual paths.

```{r data}
set.seed(42)

make_cohort <- function(n_subjects, seed = 42) {
  set.seed(seed)
  n_obs_per <- sample(6:12, n_subjects, replace = TRUE)
  do.call(rbind, lapply(seq_len(n_subjects), function(i) {
    n     <- n_obs_per[i]
    base  <- rnorm(1, mean = 7.5, sd = 1.2)
    trend <- rnorm(1, mean = -0.02, sd = 0.01)
    time  <- sort(runif(n, 0, 24))
    value <- base + trend * time + rnorm(n, sd = 0.4)
    data.frame(
      subject_id  = sprintf("P%03d", i),
      visit_month = time,
      hba1c       = pmax(4, value),
      arm         = ifelse(i <= round(n_subjects / 2), "Treatment", "Control"),
      stringsAsFactors = FALSE
    )
  }))
}

df_small <- make_cohort(25)
df_large <- make_cohort(100)
```

---

# Part I — `geom_kodom_heatmap()`

The heatmap geom bins the time axis into equal-width intervals and fills each
(subject × bin) cell with an aggregate of the `fill` aesthetic. Because every
subject occupies exactly one row regardless of visit count, the layout stays
readable even for large cohorts.

## 1. Basic usage

Map `fill` (not `colour`) to the measurement value. The stat handles
binning and aggregation automatically; the default is 10 equal-width bins
averaged by mean.

```{r heatmap-basic}
ggplot(df_small, aes(x = visit_month, id = subject_id, fill = hba1c)) +
  geom_kodom_heatmap() +
  scale_fill_kodom(name = "HbA1c (%)") +
  labs(x = "Visit (months)", y = "") +
  theme_kodom()
```

Each cell is the mean HbA1c for that subject during that time window. White
borders separate cells (controlled by `colour` and `linewidth` aesthetics on
the geom, not the scale).

---

## 2. Choosing bin resolution with `bins`

Fewer bins give a broader summary; more bins reveal finer temporal structure.
Compare 5, 10, and 20 bins on the large cohort:

```{r heatmap-bins5, fig.height = 8}
ggplot(df_large, aes(x = visit_month, id = subject_id, fill = hba1c)) +
  geom_kodom_heatmap(bins = 5, sort_by = "mean") +
  scale_fill_kodom(name = "HbA1c (%)") +
  labs(x = "Visit (months)", y = "", title = "bins = 5") +
  theme_kodom() +
  theme(axis.text.y = element_blank())
```

```{r heatmap-bins20, fig.height = 8}
ggplot(df_large, aes(x = visit_month, id = subject_id, fill = hba1c)) +
  geom_kodom_heatmap(bins = 20, sort_by = "mean") +
  scale_fill_kodom(name = "HbA1c (%)") +
  labs(x = "Visit (months)", y = "", title = "bins = 20") +
  theme_kodom() +
  theme(axis.text.y = element_blank())
```

More bins can produce empty cells (grey) when subjects miss a window entirely.
Fewer bins mask short-term fluctuations but give a cleaner read.

---

## 3. Clinical time windows with `breaks`

Instead of equal-width bins, supply explicit boundaries aligned to study
milestones. Here we use 0, 6, 12, 18, and 24 months to match typical
quarterly assessment windows.

```{r heatmap-breaks}
ggplot(df_small, aes(x = visit_month, id = subject_id, fill = hba1c)) +
  geom_kodom_heatmap(
    breaks  = c(0, 6, 12, 18, 24),
    sort_by = "mean"
  ) +
  scale_fill_kodom(name = "HbA1c (%)") +
  labs(x = "Study quarter (months)", y = "") +
  theme_kodom()
```

`breaks` overrides `bins`. Bin midpoints are used as the x-axis position for
each tile, so `scale_x_continuous()` labels remain meaningful.

---

## 4. Aggregation functions with `fun`

The default `fun = "mean"` smooths over within-window variation. Other options
surface different clinical signals:

- `"first"` / `"last"` — baseline or endpoint value in each window
- `"max"` — worst reading per window (useful for hypoglycaemia risk)
- `"min"` — best reading per window

```{r heatmap-fun-first}
ggplot(df_small, aes(x = visit_month, id = subject_id, fill = hba1c)) +
  geom_kodom_heatmap(
    breaks  = c(0, 6, 12, 18, 24),
    fun     = "first",
    sort_by = "mean"
  ) +
  scale_fill_kodom(name = "HbA1c (%)") +
  labs(x = "Study quarter", y = "", title = 'fun = "first" — earliest reading per window') +
  theme_kodom()
```

```{r heatmap-fun-max}
ggplot(df_small, aes(x = visit_month, id = subject_id, fill = hba1c)) +
  geom_kodom_heatmap(
    breaks  = c(0, 6, 12, 18, 24),
    fun     = "max",
    sort_by = "mean"
  ) +
  scale_fill_kodom(name = "HbA1c (%)") +
  labs(x = "Study quarter", y = "", title = 'fun = "max" — peak reading per window') +
  theme_kodom()
```

---

## 5. Large cohorts — where heatmap wins

`geom_kodom_line()` becomes unreadable beyond ~50 subjects; the heatmap
stays legible at 100 or more. Suppress y-axis text for large cohorts.

```{r heatmap-large, fig.height = 9}
ggplot(df_large, aes(x = visit_month, id = subject_id, fill = hba1c)) +
  geom_kodom_heatmap(bins = 12, sort_by = "mean") +
  scale_fill_kodom(
    discretize   = TRUE,
    color_breaks = c(5.7, 6.5, 8),
    name         = "HbA1c (%)"
  ) +
  labs(
    x     = "Visit (months)",
    y     = "",
    title = "100-patient cohort — discrete clinical bands"
  ) +
  theme_kodom() +
  theme(axis.text.y = element_blank())
```

`discretize = TRUE` with clinical thresholds (normal < 5.7, pre-diabetic
5.7–6.5, diabetic 6.5–8, poorly controlled > 8) converts the continuous
gradient into solid color bands. Each row becomes an instant status
classification.

---

## 6. Faceting by treatment arm

Because `geom_kodom_heatmap()` is a standard ggplot2 layer, `facet_wrap()`
works without any extra configuration. Lane ordering is computed within each
facet panel independently.

```{r heatmap-facet, fig.height = 7}
ggplot(df_large, aes(x = visit_month, id = subject_id, fill = hba1c)) +
  geom_kodom_heatmap(bins = 8, sort_by = "mean") +
  scale_fill_kodom(
    discretize   = TRUE,
    color_breaks = c(5.7, 6.5, 8),
    name         = "HbA1c (%)"
  ) +
  facet_wrap(~arm, ncol = 1, scales = "free_y") +
  labs(x = "Visit (months)", y = "", title = "Treatment vs. Control") +
  theme_kodom() +
  theme(axis.text.y = element_blank())
```

`scales = "free_y"` gives each panel its own set of lanes so subjects are
not shared across facets — each panel shows only the subjects in that arm.

---

## 7. Adjusting tile borders

The `colour` and `linewidth` aesthetics on the tile control cell borders, not
the fill scale. Pass them as fixed values directly to the geom.

```{r heatmap-borders}
ggplot(df_small, aes(x = visit_month, id = subject_id, fill = hba1c)) +
  geom_kodom_heatmap(
    breaks    = c(0, 6, 12, 18, 24),
    colour    = "grey60",
    linewidth = 0.8
  ) +
  scale_fill_kodom(name = "HbA1c (%)") +
  labs(x = "Study quarter", y = "", title = "Visible grey cell borders") +
  theme_kodom()
```

Set `colour = NA` or `linewidth = 0` to remove borders entirely for a
seamless mosaic appearance.

---

# Part II — `geom_kodom_circular()`

The circular geom projects each subject onto a radial spoke, with time
increasing outward from the centre and value encoded as a colour gradient
along the spoke. The layout resembles a Kadam flower
(*Neolamarckia cadamba*), giving the package its name.

Internally the stat converts lane ranks to angles and time to radius, then
renders in ordinary Cartesian space — no `coord_polar()` is used. Always
pair with `coord_fixed()` to preserve the circular shape.

---

## 8. Basic circular plot

Map `colour` to the measurement value, just as with `geom_kodom_line()`.
Add `coord_fixed()` and `theme_kodom_circular()`.

```{r circular-basic, fig.height = 7}
ggplot(df_small, aes(x = visit_month, id = subject_id, colour = hba1c)) +
  geom_kodom_circular() +
  scale_colour_kodom(name = "HbA1c (%)") +
  coord_fixed() +
  theme_kodom_circular() +
  labs(title = "Circular swimlane — 25 patients")
```

Each spoke is one patient. Time increases outward; teal (low HbA1c) and red
(high) are drawn as smooth gradients along each spoke.

---

## 9. Lane ordering and the visual effect on the flower

`sort_by` controls the angular order of spokes. `"mean"` groups similar
patients into arcs of the same hue, revealing structure that `"none"` scatters
randomly.

```{r circular-sort-none, fig.height = 7}
ggplot(df_small, aes(x = visit_month, id = subject_id, colour = hba1c)) +
  geom_kodom_circular(sort_by = "none") +
  scale_colour_kodom(name = "HbA1c (%)") +
  coord_fixed() +
  theme_kodom_circular() +
  labs(title = 'sort_by = "none" — random spoke order')
```

```{r circular-sort-mean, fig.height = 7}
ggplot(df_small, aes(x = visit_month, id = subject_id, colour = hba1c)) +
  geom_kodom_circular(sort_by = "mean") +
  scale_colour_kodom(name = "HbA1c (%)") +
  coord_fixed() +
  theme_kodom_circular() +
  labs(title = 'sort_by = "mean" — high-mean spokes cluster together')
```

---

## 10. The seam gap with `gap_fraction`

`gap_fraction` (default `0.15`) leaves an empty wedge at the ordering seam
so the first and last subject are clearly separated. Increase it for a
fan/semicircle layout; reduce it toward `0` for a nearly complete ring.

```{r circular-gap-small, fig.height = 7}
ggplot(df_small, aes(x = visit_month, id = subject_id, colour = hba1c)) +
  geom_kodom_circular(sort_by = "mean", gap_fraction = 0.03) +
  scale_colour_kodom(name = "HbA1c (%)") +
  coord_fixed() +
  theme_kodom_circular() +
  labs(title = "gap_fraction = 0.03 — near-complete ring")
```

```{r circular-gap-large, fig.height = 7}
ggplot(df_small, aes(x = visit_month, id = subject_id, colour = hba1c)) +
  geom_kodom_circular(sort_by = "mean", gap_fraction = 0.35) +
  scale_colour_kodom(name = "HbA1c (%)") +
  coord_fixed() +
  theme_kodom_circular() +
  labs(title = "gap_fraction = 0.35 — fan layout")
```

---

## 11. Hollow centre with `inner_fraction`

`inner_fraction` (default `0.3`) adds a radial buffer so that subjects with
short follow-up are not collapsed to a dot at the origin. Increase it for
more whitespace; set it to `0` to start spokes at the centre.

```{r circular-inner-small, fig.height = 7}
ggplot(df_small, aes(x = visit_month, id = subject_id, colour = hba1c)) +
  geom_kodom_circular(sort_by = "mean", inner_fraction = 0) +
  scale_colour_kodom(name = "HbA1c (%)") +
  coord_fixed() +
  theme_kodom_circular() +
  labs(title = "inner_fraction = 0 — spokes from the centre")
```

```{r circular-inner-large, fig.height = 7}
ggplot(df_small, aes(x = visit_month, id = subject_id, colour = hba1c)) +
  geom_kodom_circular(sort_by = "mean", inner_fraction = 0.6) +
  scale_colour_kodom(name = "HbA1c (%)") +
  coord_fixed() +
  theme_kodom_circular() +
  labs(title = "inner_fraction = 0.6 — wide hollow centre")
```

---

## 12. Clockwise vs. counter-clockwise with `direction`

`direction = 1L` (default) places lanes clockwise from the top.
`direction = -1L` reverses the ordering.

```{r circular-ccw, fig.height = 7}
ggplot(df_small, aes(x = visit_month, id = subject_id, colour = hba1c)) +
  geom_kodom_circular(sort_by = "mean", direction = -1L) +
  scale_colour_kodom(name = "HbA1c (%)") +
  coord_fixed() +
  theme_kodom_circular() +
  labs(title = "direction = -1L — counter-clockwise")
```

---

## 13. Suppressing or resizing observation points

`show_points = FALSE` removes the markers, leaving only the gradient path.
`size` controls point size exactly as in `geom_kodom_line()`. A smaller
default (`size = 2.0`) suits the compact circular layout.

```{r circular-no-points, fig.height = 7}
ggplot(df_small, aes(x = visit_month, id = subject_id, colour = hba1c)) +
  geom_kodom_circular(sort_by = "mean", show_points = FALSE) +
  scale_colour_kodom(name = "HbA1c (%)") +
  coord_fixed() +
  theme_kodom_circular() +
  labs(title = "show_points = FALSE — gradient paths only")
```

```{r circular-big-points, fig.height = 7}
ggplot(df_small, aes(x = visit_month, id = subject_id, colour = hba1c)) +
  geom_kodom_circular(sort_by = "mean", size = 3.5, linewidth = 0.3) +
  scale_colour_kodom(name = "HbA1c (%)") +
  coord_fixed() +
  theme_kodom_circular() +
  labs(title = "Larger points, thinner paths")
```

---

## 14. Large cohort — circular as a population fingerprint

With 100 subjects the individual spokes pack tightly, creating a dense ring
that functions as a population-level visual fingerprint. Suppressing points
and using thin paths keeps the display legible.

```{r circular-large, fig.height = 8}
ggplot(df_large, aes(x = visit_month, id = subject_id, colour = hba1c)) +
  geom_kodom_circular(
    sort_by     = "mean",
    show_points = FALSE,
    linewidth   = 0.4,
    gap_fraction = 0.08
  ) +
  scale_colour_kodom(
    discretize   = TRUE,
    color_breaks = c(5.7, 6.5, 8),
    name         = "HbA1c (%)"
  ) +
  coord_fixed() +
  theme_kodom_circular() +
  labs(title = "100 patients — population fingerprint")
```

Subjects with high mean HbA1c cluster in one arc; well-controlled patients
occupy the opposite arc. The colour ring makes the population split visible
at a glance.

---

## 15. Combining circular and heatmap: a side-by-side summary

The two layouts are complementary. Use `patchwork` (or `gridExtra`) to place
them together for a publication-ready comparison panel.

```{r combined, fig.height = 6, message = FALSE}
library(patchwork)

p_heat <- ggplot(df_small, aes(x = visit_month, id = subject_id, fill = hba1c)) +
  geom_kodom_heatmap(bins = 6, sort_by = "mean") +
  scale_fill_kodom(name = "HbA1c (%)") +
  labs(x = "Visit (months)", y = "", title = "Heatmap") +
  theme_kodom(legend_position = "bottom")

p_circ <- ggplot(df_small, aes(x = visit_month, id = subject_id, colour = hba1c)) +
  geom_kodom_circular(sort_by = "mean", show_points = FALSE) +
  scale_colour_kodom(name = "HbA1c (%)") +
  coord_fixed() +
  theme_kodom_circular() +
  labs(title = "Circular")

p_heat + p_circ
```
